Metric Spaces with Expensive Distances

### 2019-04-23 paper/expensive-distances-main.tex
almost final

\section{Background and Definitions}
%
\myparagraph{Doubling dimension}
A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
and $k$ is the smallest number having that property.
The \textit{doubling dimension} of a doubling space is defined as $\log k$
(since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
we always use $\log$ to denote the logarithm with base 2).
It is easy to see that a subspace of a space with doubling dimension $d$ 
is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).

@@

On the other hand, spanner constructions based on WSPD\footnote{Well-Separated Pair Decomposition; 
for convenience we remind the definition in \cref{sec:appendix_background}} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

@@

From these experiments we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is significantly worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners.

The persistence diagrams that we used as
realistic data sets,
come from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only used
dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with \libraryname{DIPHA}
library.

Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
\grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
any advantage of our algorithm disappears. However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
We can offer the following explanation, assuming the reader is familiar
with the definition of $\wsdist{q}$. For higher values of $q$,
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, then we
reduce the number of distances computed by a factor of 10.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in brute-force column
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, as done in \dionisus, or
it can be computed with the user-given accuracy using the auction algorithm,
as in \hera. Exact computation is very expensive (see \cite{kerber2017geometry}), and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
In order to make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.

@@

The proof idea is the following. First, we notice
that current candidate $c$ allows us to exclude points
$p_i$ that are too far from $c$. This means that if $p_i$ is
outside of the ball of radius $2\dist(c, q)$ centered at $c$, then the lower
bound for $p_i$ will be larger than $\dist(c, q)$,
and discarding $p_i$ does not cost us a distance computation.
On the other hand, once we computed a distance to $c$
from a point $p_j$, which is inside this ball but happens to be farther from $q$ than $c$,
we can use it to discard points which are very close to $p_j$, 
namely, inside a small ball of radius approximately $\eps \dist(c, q)$, because we
are solving approximate nearest neighbour problem.
Then we use a standard packing argument for doubling spaces, applied to the large and small balls, to show
that the number of distance computations that we pay per candidate is constant.
It remains only to prove that the number of different candidates is $O(\log n)$
in expectation; there is a technicality that prevents using the standard
backwards analysis argument verbatim, but it can be adapted.
The full proof is given in \cref{sec:proof_ann}.

@@

\begin{figure}[hb!]
    \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
    \caption{Example of point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
    The exact nearest neighbour is the point $p_1$, next point $p_i$ is placed 
    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
    is the true nearest neighbour cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
    \label{fig:exact_bad_example}
\end{figure}

@@

\myparagraph{Wasserstein Distance}
A persistence diagram is a multiset of points $(x, y)$ above the diagonal
$\Delta$ and all points of $\Delta$ with infinite cardinality.

>>>>

\section{Background}
%
\myparagraph{Doubling dimension}
A metric space is called \textit{doubling} with \textit{doubling constant} $k$,
if every ball of radius $r$ can be covered by at most $k$ balls of radius $r/2$,
and $k$ is the smallest number having that property.
The \textit{doubling dimension} of a doubling space is defined as $\log k$
(since we usually ignore multiplicative constants, the base of the logarithm is not really important; however,
we always use $\log$ to denote the logarithm with base 2).
It is easy to see that a subspace of a space with doubling dimension $d$ 
is always doubling and has the doubling dimension $O(d)$ (but not necessarily $d$).

@@

On the other hand, spanner constructions based on Well-Separated Pair Decomposition\footnote{ 
for convenience, we remind the definition of WSPD in \cref{sec:appendix_background}} only compute
$O(n\log n+n \eps^{-d})$ distances 
to construct a \espanner in doubling dimension $d$.
The spanner size is $O(n\eps^{-d})$. Assuming $\eps$ and $d$ again as constants,
this construction yields a $O(n)$-size spanner using only $O(n\log n)$ distance
computations. However, the algorithm is significantly more involved.

@@

From these experiments we conclude that the algorithm with the highest chance
of saving a significant amount of distance computations on real data is \bgrdy. 
Even though \brndm also substantially reduces
the number of computed distances, its performance is worse;
other algorithms that we tested do not produce sparse spanners.
While it would seem plausible that access to approximate value of the distance
could be exploited in the spanner construction,
we could not find a working heuristic; \bqsgrdy and \bqsshaker produce extremely
dense spanners. More data from experiments in $\RR^n$ can be found in
\cref{sec:spanners_euclidean_addendum}.

The realistic data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only used
dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with \libraryname{DIPHA}
library \cite{bauer2014distributed}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
\grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
any advantage of our algorithm disappears. However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
We can offer the following explanation\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}. For higher values of $q$,
\label{}
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, then we
reduce the number of distances computed by a factor of 10.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in the brute-force columns
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, or
it can be computed with the user-given accuracy with the auction algorithm,
as in \hera. Exact computation is very expensive \cite{kerber2017geometry}, and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
To make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.

@@

The proof idea is the following. First, we notice
that current candidate $c$ allows us to exclude points
$p_i$ that are too far from $c$. This means that if $p_i$ is
outside of the ball of radius $2\dist(c, q)$ centered at $c$, then the lower
bound for $p_i$ will be larger than $\dist(c, q)$,
and discarding $p_i$ does not cost us a distance computation.
On the other hand, once we computed a distance to $c$
from a point $p_j$, which is inside this ball but happens to be farther from $q$ than $c$,
we can use it to discard points which are very close to $p_j$, 
namely, inside a small ball of radius approximately $\eps \dist(c, q)$, because we
are solving approximate nearest neighbour problem.
Then we use a standard packing argument for doubling spaces, applied to the large and small balls, to show
that the number of distance computations that we pay per candidate is constant.
It remains only to prove that the number of different candidates is $O(\log n)$
in expectation; the standard
backwards analysis argument cannot be applied verbatim, because of a technicality,
but it can be adapted.
The full proof is given in \cref{sec:proof_ann}.

@@

\begin{figure}[hb!]
    \includegraphics[width=0.7\textwidth]{\expDistFigPath exact_bad_example.png}
    \caption{Example of a point set where exact nearest neighbour search cannot be accelerated by maintaining bounds.
    The exact nearest neighbour is the point $p_1$, next point $p_i$ is placed 
    in the curvilinear triange formed by the balls around the query point, $p_2$ and $p_{i-1}$. Even verifying that $p_1$
    is the true nearest neighbour cannot be done without computing all distances $\dist(p_i,q)$. Indeed, every computed
    $\dist(p_i,q)$ allows to exclude the region in the corresponding ball around $p_i$, but all these balls contain only one $p_i$.}
    \label{fig:exact_bad_example}
\end{figure}

@@

\myparagraph{Wasserstein Distance}
Let $\diag$ denote the line in the plane with the equation $y = x$; we call it the diagonal.
A persistence diagram is a finite multiset of points $(x, y)$ with $x > y$
$\Delta$ and all points of $\Delta$ with infinite cardinality.
The points with $y > x$ are called \textit{off-diagonal} points,
and points on $\Delta$ are called \textit{diagonal} points.
The \emph{$q$-Wasserstein distance} is defined as
\[
    \wsdist{q}(X,Y) = \left[ \inf_{\eta:X\rightarrow Y} \sum_{x\in X} \|x-\eta(x)\|_\infty^q\right]^{1/q},
\]
where $\eta$ ranges over all bijections between $X$ and $Y$.
Equivalent definition, which is better suited for actual computations,
is as follows. Let $\pi\colon \RR^2 \to \Delta$ be defined
as $(x,y) \mapsto (\frac{x+y}{2}, \frac{x+y}{2})$; we refer to $\pi$ as \textit{diagonal
projection}. Let $\tX$ be the set $(X\setminus \Delta) \cup \pi(Y \setminus \Delta)$
and $\tY$ be the set $(Y\setminus \Delta) \cup \pi(X \setminus \Delta)$; in plain words,
to get $\tX$, we add to the off-diagonal points of $X$
the diagonal projections of the off-diagonal points of $Y$,
and perform the same operation to get $\tY$. Notice that $\tX$ and $\tY$
are finite sets of the same cardinality. Let $G$ be a complete bipartite
graph with vertices $\tX \coprod \tY$, where each vertex from $\tX$
is connected with each vertex of from $\tY$. If edge $(x,y)$ of $G$
has at least one of the vertices $x$ or $y$ not on the diagonal,
then the edge is assigned weight $\ell_{\infty}(x, y)^q$, otherwise
the edge is assigned weight 0. Wasserstein distance is defined as the $q$-th
root of the minimal cost of a perfect matching on $G$; in other words,
in order to find $\wsdist{q}$, one must solve an \textit{assignment problem}.
